---
title: "Functions"
author: "Richard G. Gardiner"
date: "May 20, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions

All R functions have three parts:

- `body()`: the code inside the function
- `formals()`: the list of arguments which controls how you can call the function.
- `environment()`: the "map" of hte location of the function's variables.

You can have R show you each of these:

```{r}
f <- function(x) {
  x^2
}

f
```

```{r}
formals(f)
```

```{r}
body(f)
```

```{r}
environment(f)
```

The one exception to the three parts rule are primitative functions like `sum()` which call C code directly and contain no R code.  There `formals()`, `body()`, and `environment()` are all thus NULL. Primitive functiosn are only found in the `base` package, and since they operate at a low level, they can b more efficient and can have different rules for argument matching.  This, however, comes at a cost of behaving differently from all other functions in R.  Hence the R core team generally avoids creating them unless there is no other option.



