---
title: "Chapter 1: Data Structures"
author: "Richard G. Gardiner"
date: "1/23/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

There are five main types of data structures that can be organized by dimensionality and whether they're homogenous (all contents must be the same type) or heterogreneeous (the contents can be of different types):

```{r}
library(tidyverse)
homogenous <- c("Atomic Vector", "Matrix", "Array")
heterogeneous <- c("List", "Data Frame", "na")

data_types <- cbind(homogenous, heterogeneous) 
row.names(data_types) <- c("1d", "2d", "Nd")
data_types <- as.data.frame(data_types)

data_types
```

Almost all other objects are built upon these foundations.  The best way to understand the data structures of an object is to use the `str()` command.  This gives a human readable, and descriptive explanation of any R data structure.


## Vectors

The basic data structure in R is the vector. Vectors come in two types: atomic vectors and lists.  They have three common properites:

- Type, `typeof()`, what it is.
- Length, `length()`, how many elements it contains.
- Attributes, `attributes()`, additional arbitrary metadata.

All elements of an atomic vector must be the same type, whereas the elements of a list can have different types.  If you want to know if it is a vector, use the following code: `is.atomic(x) || is.list(x)`.


### Atomic Vectors

There are four common types of atomic vectors: logical, integer, double (often called numeric) and character.  There are two very rare types that aren't discussed in the book: complex and raw.

Atomic vectors are usually created with `c()`, short for combine:

```{r}
dbl_var <- c(1, 2.5, 4.5)

# with the L suffix, you get an integer rather than a double
int_var <- c(1L, 6L, 10L)

# Use TRUE and FALSE (or T and F) to create logical vectors
log_var <- c(TRUE, FALSE, T, F)

char_var <- c("These are", "some strings")
```

Atomic vectors are always flat, even if you nest `c()`'s:

```{r}
c(1, c(2, c(3, 4)))

# the same as:
c(1, 2, 3, 4)
```


#### Types  and tests

Given a vector, you can determine its type with `typeof()`, or check if its a sepcific type with an "is" function: `is.character()`, `is.double()`, `is.integer()`, `is.logical()`, or, more generally, `is.atomic()`.

```{r}
typeof(int_var)

is.integer(int_var)

is.atomic(int_var)

typeof(dbl_var)

is.double(dbl_var)

is.atomic(dbl_var)
```

The `is.numeric()` function is a general test for the "numberliness" of a vector and returns `TRUE` for both integers and double vectors.  It is not a specific test for double vectors, which are often called numberic:

```{r}
is.numeric(int_var)
is.numeric(dbl_var)
```

#### Coercion

All elemtns of an atomic vector must be the same type, so when you attempt to combine different types they will be **coerced** to the most flexible type.  Types from least to move flexible are: logical, integer, double, and character.

For example, combining a character and an integer yields a character:

```{r}
str(c("a", 1))
```

When a logicla vector is coerced to an integer or double, `TRUE` becomes 1 and `FALSE` becomes 0.  THis is very useful in conjunction with `sum()` and `mean()`.

```{r}
x <- c(FALSE, FALSE, TRUE)

as.numeric(x)

# Total number of TRUEs
sum(x)

# Proprotion that are TRUE
mean(x)
```

Coercion often happens automatically. Most mathematical functions (+, log, abs, etc.) will coerce to a double or integer, and most logical operations (&, |, any, etc) will coerce to a logical. You will usually get a warning message if the coercion might lose information. If confusion is likely, explicitly coerce with `as.character()`, `as.double()`, `as.integer()`, or `as.logical()`.



