---
title: 'Chatper 3: Subsetting Data'
author: "Richard G. Gardiner"
date: "April 3, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Atomic Vectors

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

There are five things that you cna use to subset a vector:

1. Positive integers return elements at the specified positions:

```{r}
x[c(3, 1)] # returns items in position 3 and position 1
x[c(1, 1)] # returns the first item 2 times
```

2. Negative integers omit elments at the specified positions:


```{r}
x[-c(3, 1)] # includes everything but the numbers at position 3 and 1
```

3. Logical vectors select elements where the corresponding logical value is true.  THis is probably the most useful type of subsetting because you write the expression that creates a logical vector:

```{r}
x[c(TRUE, TRUE, FALSE, FALSE)]

x[x > 3]
```

If the logical vector is shorter than the vector being subsetted, it will be recycled to be the same length:

```{r}
x[c(TRUE, FALSE)]
x[c(TRUE, FALSE, TRUE, FALSE)]
```

4. Nothing returns to original vector.  Not very helpful for vectors but good for other types:

```{r}
x[]
```

5. Zero returns a zero length vector.  It is not something you do on purpose, but it can be helpful for generating test data.

```{r}
x[0]
```

